Approximating the matrix exponential derivative with respect to the CTMC generator matrix, with application to global SARS-CoV-2

### 2023-04-05 manuscript/main.tex
abstract

\title{Approximating the matrix exponential derivative of the CTMC generator matrix, with application to global SARS-CoV-2}

%\date{}

@@

We investigate the accuracy of a first-order approximation to the directional derivative of the matrix exponential evaluated on continuous-time Markov chain (CTMC) generator matrices.

>>>>

\title{Approximating the matrix exponential derivative with respect to the CTMC generator matrix, with application to global SARS-CoV-2}

\date{}

@@

The continuous-time Markov chain (CTMC) is the workhorse model of evolutionary biology.  Learning CTMC model parameters using modern, gradient-based methods requires the derivative of the matrix exponential evaluated at the CTMC's infinitesimal generator (rate) matrix.  Motivated by the derivative's extreme computational complexity as a function of state space cardinality, recent work demonstrates the surprising effectiveness of a fast, first-order approximation for a host of problems in computational biology.  We obtain rigorous bounds for the error accrued by such an approach and establish a ``blessing of dimensionality'' result stemming from the novel insight that, in a certain sense, high-dimensional CTMC quickly converge to stationarity.  Finally, we apply the first-order approximation within surrogate-trajectory Hamiltonian Monte Carlo to the analysis of the early spread of SARS-CoV-2 across 44 geographic regions that comprise a state space of unprecedented dimensionality within evolutionary biology.
